Partitioned Local Depth (PaLD) Clustering Analyses in R

### 2022-05-22 r-journal/manuscript.Rmd
update with Ken's comments

Building on existing approaches to (global) depth, local depth expresses features of centrality as an interpretable probability which is free of parameters and robust to outliers.  Then, partitioning the probability which defines local depth, we obtain a measure of cohesion between pairs of points.  Both local depth and cohesion reflect aspects of relative position (rather than absolute distance) and provide a straightforward way to account for varying density across the space.  Specifically, provided that two sets are separated (in the sense that the minimum between-set distance is greater than the maximum within-set distance), cohesion is invariant under the contraction and dilation of the distances within each set.  This property may be particularly valuable when one has reason to believe that the magnitude of distances varies across the space.

@@

As seen throughout this paper, we display the network obtained from cohesion using a force-directed graph drawing algorithm and emphasize the strong ties (colored by connected component).  We refer to the connected components of the network of strong ties as "clusters."  Note that to qualify as a cluster in this definition, one may not have any strong ties with those outside the cluster, and thus the existence of disjoint groups is a strong signal for separation.  Here, clusters are identified without additional user inputs nor optimization criteria.  If one wishes to further break the community graph into groups, one may use community detection methods for networks (such as spectral clustering or the Louvain algorithm).  Though only briefly considered here, one may also use the collection of strong ties in place of (weighted) k-nearest neighbors in settings such as classification and smoothing.  Overall, the structural information obtained from local depth, cohesion and community graphs can provide a holistic perspective on the data which does not require the use of distributional assumptions, optimization criteria nor additional user inputs.

We present a new package, \CRANpkg{pald}, for calculating Partitioned Local Depths (PaLD) probabilities, implementing clustering analyses, and creating data visualizations to represent the clusters. This paper will describe how to use the package as well as walk through two examples.

@@

1. Helper functions to organize data into the correct format, into distance matrices and then cohesion matrices
2. Functions that convert a cohesion matrix into a variety of useful formats, including partitioned local depths, clusters, and graphs
3. Plotting functions

@@

For demonstration purposes, below is a sample data frame with two variables, `x1` and `x2`. The methods put forth here work on data frames with higher dimensions, as described in the **Examples** section; we are simply choosing a small data frame here for demonstration purposes.

```{r}
library(pald)
df <- data.frame(
  x1 = c(6, 8, 11, 16, 4, 14),
  x2 = c(5, 4, 13, 7, 6, 10)
)
rownames(df) <- c("A", "B", "C", "D", "E", "F")
```

The first step needed to calculate the partitioned local depths is to construct a _distance matrix_. If the data are already in this form, the user can skip to the next step. The `dist()` function converts an input data frame into a distance matrix, as demonstrated below.

@@

This will create a `dist` object. If converted to a matrix, this will be a $n\times n$ distance matrix, where $n$ corresponds to the number of observations in the original data frame, in this example $n = 6$.

This dist object, or equivalently a distance matrix, can then be passed to the `cohesion_matrix()` function in order to calculate the pairwise cohesion values. Cohesion is an interpretable probability that reflects the strength of alignment of two points. Again, if the user begins with a distance matrix, they can skip the first step and simply input the distance matrix into this function.

@@

The _cohesion matrix_ output by the `cohesion_matrix()` is the main input for the majority of the remaining functions.

### Functions that convert a cohesion matrix into useful formats

@@

To calculate the _clusters_ that each point will fall into, we can use the `community_clusters()` function. This will output a data frame with two columns, the first will correspond to the `point`, as identified by the row name of the original input data frame, `df`, the second will identify the `cluster` that each point belongs to.

@@

The `strong_threshold()` function will calculate the cohesion threshold for strong ties. This is equal to half the average of the diagonal cohesion matrix. This is a threshold that may be used to distinguish between strong and weak ties.

@@

In this case, the threshold is `0.12`.

@@

The function `cohesion_strong()` will update the cohesion matrix to set all weak ties to zero (via the `strong_threshold()` function). Optionally, the matrix will also be symmetrized, with the default parameter `symmetric = TRUE`.

Finally, the `community_graphs()` function takes the cohesion matrix and creates \CRANpkg{igraph} objects, graphs that describes the relationship between the points. This function will output a list of three objects:

Here we see that there are three connected components, points `A` and `B`, and `A` and `E` which form the first cluster, and points `D` and `F` which form another.

@@

The final category of function is functions for data visualization. We can begin by visualizing the points in data frame `df` (Figure \ref{fig:fig1}). When visualizing these points, it is important to have a square plot so as to not distort the distances. When using the \CRANpkg{ggplot2} package for this visualization, you can use the `aspect.ratio  = 1` argument in the `theme()` function. If using the `plot()` function included in the base library, you can set the plot area to be a square by using `par(pty = "s")` prior to calling the `plot()` function, as shown below.

```{r fig1, fig.cap = "Visualize the points from data frame `df`", message = FALSE, warning = FALSE}
library(ggplot2)
ggplot(df, aes(x1, x2)) +
  geom_text(label = rownames(df)) + 
  theme(aspect.ratio = 1)
```

@@

The `layout` argument allows the user to pass a matrix to dictate the layout of the graph. For example, if we wanted the graph to match the visualization displayed in Figure \ref{fig:fig1}, we can pass `as.matrix(df)`, or a matrix of the data frame `df` to the `layout` argument  (Figure \ref{fig:fig3}.  Notice in the code below, we set `par(pty = "s")` in order to ensure that we have a square plotting region. Additionally, this `plot_community_graphs()` function will also permit parameters that can be passed to the `plot.igraph()` function. For example, we can pass arguments to the `plot.igraph` function via the `...`, for example to increase the vertex size and change the vertex label color, we can specify `vertex.size = 15` and `vertex.label.color = "white"`.

```{r fig3, fig.cap = "PaLD graph displaying the relationship between the points in data frame `d`, matching the original layout in Figure 1"}
par(pty = "s")

plot_community_graphs(cohesion, 
                      layout = as.matrix(df),
                      vertex.size = 15,
                      vertex.label.color = "white")
```

@@

The first example if from a subset of tissue gene expression data from \citet{zilliox2007gene}, \citet{mccall2011gene}, and \citet{mccall2014gene}, obtained from the **tissuesGeneExpression** bioconductor package \citep{tissue}. A `dist` object was created using this data set and is included the \CRANpkg{pald} package in an object called `tissue_dist`.

@@

We can also display the relationships between tissue samples using the `plot_community_graphs()` function (Figure \ref{fig:fig4}). For clarity of the display, we show how to remove the labels using `show_labels = FALSE`. We will instead color by the labels by passing these to the `vertex.color` parameter through the  `...` to the `plot.igraph` function. Similarly, we can add a legend using the `legend()` function, as you would for a \CRANpkg{igraph} visualization. Additionally, we use the `edge_width_factor` and `emph_strong` arguments to adjust the width of the lines between and within PaLD clusters.

@@

```{r fig4, echo = FALSE, message = FALSE, warning = FALSE, fig.cap = "PaLD clustering of tissue data. The line colors indicate the PaLD clusters, the point colors indicate the tissue classification.", out.width = "100%"}
knitr::include_graphics("fig5.png")
```

@@

We can then use the `neighbors()` function from the \CRANpkg{igraph} package to extract the neighbors in this graph. For example, if we wanted to extract all neighbors where the language is "French", we would run the following.

@@

The \CRANpkg{pald} includes three randomly generated data frames corresponding to plots from @berenhaut2022social:

@@

We will demonstrate how we can compare PaLD to two clustering methods: *k*-means and hierarchical clustering. The code below calculates the cohesion matrix (`exdata_cohesion`) as well as the clusters via PaLD (`exdata_pald`), *k*-means (`exdata_kmeans`) and hierarchical clustering (`exdata_hclust`).

@@

```{r fig5, fig.cap = "PaLD clustering of randomly generated example data (from Figure 4D from Berenhaut et al. (2022)) compared to *k*-means and hierarchical clustering with k = 8."}
par(mfrow = c(1, 3), pty = "s")
plot(
  exdata3,
  pch = 16,
  col = pald_colors[exdata_pald],
  xlab = "",
  ylab = "",
  main = "PaLD Clusters"
)
plot(
  exdata3,
  pch = 16,
  col = pald_colors[exdata_kmeans],
  xlab = "",
  ylab = "",
  main = "K-Means Clusters (k = 8)"
)
plot(
  exdata3,
  pch = 16,
  col = pald_colors[exdata_hclust],
  xlab = "",
  ylab = "",
  main = "Hiearchical Clusters (k = 8)"
)
```

@@

The ability of the PaLD algorithm to discern clusters is demonstrated here.

@@

This paper introduces the \CRANpkg{pald} package, demonstrating it's utility for providing a parameter-free clustering algorithm that can easily be applied to any data set.

>>>>

Building on existing approaches to (global) depth, local depth expresses features of centrality as an interpretable probability which is free of parameters and robust to outliers.  Then, partitioning the probability which defines local depth, we obtain a measure of cohesion between pairs of points.  Both local depth and cohesion reflect aspects of relative position (rather than absolute distance) and provide a straightforward way to account for varying density across the space. Specifically, as shown in \citet{berenhaut2022social}, provided that two sets are separated (in the sense that the minimum between-set distance is greater than the maximum within-set distance), cohesion is invariant under the contraction and dilation of the distances within each set. This property may be particularly valuable when one has reason to believe that the magnitude of distances varies across the space.

@@

As seen throughout this paper, we display the network obtained from cohesion using a force-directed graph drawing algorithm and emphasize the strong ties (colored by connected component).  We refer to the connected components of the network of strong ties as community "clusters."  Note that to qualify as a cluster in this definition, one may not have any strong ties with those outside the cluster, and thus the existence of disjoint groups is a strong signal for separation.  Here, clusters are identified without additional user inputs nor optimization criteria.  If one wishes to further break the community graph into groups, one may use community detection methods for networks (such as spectral clustering or the Louvain algorithm).  Though only briefly considered here, one may also use the collection of strong ties in place of (weighted) k-nearest neighbors in settings such as classification and smoothing.  Overall, the structural information obtained from local depth, cohesion and community graphs can provide a holistic perspective on the data which does not require the use of distributional assumptions, optimization criteria nor additional user inputs.

We present a new package, \CRANpkg{pald}, for calculating Partitioned Local Depths (PaLD) probabilities, implementing clustering analyses, and creating data visualizations to display community structure. This paper will describe how to use the package as well as walk through two examples.

@@

1. A function for computing the cohesion matrix
2. Functions for extracting useful information from the cohesion matrix, such as local depths, neighbors, clusters, and graph objects
3. Plotting functions for community graphs

@@

The input for the Partitioned Local Depths (PaLD) is a distance matrix or `dist` object.  Note that the collection of input distances (or dissimilarities) does not need to satisfy the triangle inequality.

For demonstration purposes, we will show how one can compute a distance matrix from an input data frame with, say, two variables `x1` and `x2`. The input data may be of any dimension. The methods put forth here work on data frames with higher dimensions, as described in the **Examples** section; we are simply choosing a small data frame here for demonstration purposes.

```{r}
library(pald)
df <- data.frame(
  x1 = c(6, 8, 8, 16, 4, 14),
  x2 = c(5, 4, 10, 8, 4, 10)
)
rownames(df) <- c("A", "B", "C", "D", "E", "F")
```

The `dist()` function converts an input data frame into a distance matrix, as demonstrated below. If the data are already provided as a distance matrix (or `dist` object), the user can skip to the next step.

@@

The function above creates a `dist` object. If converted to a matrix, this will be a $n\times n$ distance matrix, where $n$ corresponds to the number of observations in the original data frame, in this example $n = 6$.

This `dist` object, or a distance matrix, can then be passed to the `cohesion_matrix()` function in order to calculate the pairwise cohesion values.

Cohesion is an interpretable probability that reflects the strength of alignment of two points within local regions. It captures aspects of the relative positioning of points and accounts for varying density across the space.

@@

The _cohesion matrix_ output by the `cohesion_matrix()` function is the main input for the majority of the remaining functions.

### Functions for extracting information from the cohesion matrix

@@

@@

The `strong_threshold()` function will calculate the cohesion threshold for strong ties. This is equal to half the average of the diagonal cohesion matrix.\citep{berenhaut2022social} This is a threshold that may be used to distinguish between strong and weak ties.

@@

In this case, the threshold is `0.117`.

The function `cohesion_strong()` will update the cohesion matrix to set all weak ties to zero (via the `strong_threshold()` function). Optionally, the matrix will also be symmetrized, using the entry-wise (parallel) minimum of the cohesion matrix and its transpose, with the default parameter `symmetric = TRUE`.

The `community_graphs()` function takes the cohesion matrix and creates \CRANpkg{igraph} objects, graphs that describe the relationship between the points. This function will output a list of three objects:

Here we see that there are three connected components, ties `A-B` and  and `A-E` which form the first cluster, and the tie `D-F` which forms another.

@@

The "clusters" identified by PaLD are the connected components of the graph of strong ties, `G_strong`. To directly calculate them, we can use the `community_clusters()` function. This will output a data frame with two columns, the first will correspond to the `point`, as identified by the row name of the original input data frame, `df`, the second will identify the `cluster` that each point belongs to.

@@

The final category of function is functions for data visualization. We can begin by visualizing the points in data frame `df` (Figure \ref{fig:fig1}). When visualizing these points, it is important to have the aspect ratio of the x and y axes equal to 1 so as to not distort the distances. When using the \CRANpkg{ggplot2} package for this visualization, you can use the `coord_fixed(ratio = 1)` function. If using the `plot()` function included in the base library, you can use the `asp = 1` argument.

```{r fig1, fig.cap = "Visualize the points from data frame `df`", message = FALSE, warning = FALSE}
library(ggplot2)
ggplot(df, aes(x1, x2)) +
  geom_text(label = rownames(df)) + 
  coord_fixed(ratio = 1) + 
  xlim(c(4, 16)) + 
  ylim(c(4, 16))
```

@@

The `layout` argument allows the user to pass a matrix to dictate the 2-dimensional layout of the graph. For example, if we wanted the graph to match the visualization displayed in Figure \ref{fig:fig1}, we can pass `as.matrix(df)`, or a matrix of the data frame `df` to the `layout` argument  (Figure \ref{fig:fig3}). Additionally, this `plot_community_graphs()` function will also permit parameters that can be passed to the `plot.igraph()` function. For example, we can pass arguments to the `plot.igraph` function via `...`, for example to increase the vertex size and change the vertex label color, we can specify `vertex.size = 100` and `vertex.label.color = "white"`. Additionally, to allow axes, we use `axes = TRUE` and to put them back on the original scale we set `rescale = FALSE`, resetting the axis limits using `xlim` and `ylim`. The `par(pty = "s")` function forces the subsequent plot to be square.

```{r fig3, fig.cap = "PaLD graph displaying the relationship between the points in data frame `df`, matching the original layout in Figure 1"}
par(pty = "s")

plot_community_graphs(cohesion, 
                      layout = as.matrix(df),
                      vertex.size = 100,
                      vertex.label.color = "white",
                      axes = TRUE,
                      rescale = FALSE,
                      asp = 1,
                      xlim = c(4, 16),
                      ylim = c(4, 16))
```

@@

The first example is from a subset of tissue gene expression data from \citet{zilliox2007gene}, \citet{mccall2011gene}, and \citet{mccall2014gene}, obtained from the **tissuesGeneExpression** bioconductor package \citep{tissue}. A `dist` object was created using this data set and is included the \CRANpkg{pald} package in an object called `tissue_dist`.

@@

We can also display the relationships between tissue samples using the `plot_community_graphs()` function (Figure \ref{fig:fig4}). For clarity of the display, we show how to remove the labels using `show_labels = FALSE`. We will instead color by the labels by passing these to the `vertex.color` parameter through the  `...` to the `plot.igraph` function. Similarly, we can add a legend using the `legend()` function, as you would for an \CRANpkg{igraph} visualization. Additionally, we use the `edge_width_factor` and `emph_strong` arguments to adjust the width of the lines between and within PaLD clusters.

@@

```{r fig4, echo = FALSE, message = FALSE, warning = FALSE, fig.cap = "Community cluster network for the tissue data. The line colors indicate the PaLD clusters, the point colors indicate the tissue classification.", out.width = "100%"}
knitr::include_graphics("fig5.png")
```

@@

We can then use the `neighbors()` function from the \CRANpkg{igraph} package to extract the strong neighbors in this graph. For example, if we wanted to extract all neighbors where the language is "French", we would run the following.

@@

The \CRANpkg{pald} package includes three randomly generated data frames corresponding to plots from @berenhaut2022social:

@@

We will demonstrate how we can compare PaLD to two clustering methods: *k*-means and hierarchical clustering. The code below calculates the cohesion matrix (`exdata_cohesion`) as well as the clusters via PaLD (`exdata_pald`), *k*-means (`exdata_kmeans`) and hierarchical clustering using complete linkage (`exdata_hclust`).

@@

```{r fig5, fig.cap = "PaLD clustering of randomly generated example data (from Figure 4D from Berenhaut et al. (2022)) compared to k-means and hierarchical clustering with k = 8."}
par(mfrow = c(1, 3), pty = "s")
plot(
  exdata3,
  pch = 16,
  col = pald_colors[exdata_pald],
  xlab = "",
  ylab = "",
  main = "PaLD Clusters",
  asp = 1
)
plot(
  exdata3,
  pch = 16,
  col = pald_colors[exdata_kmeans],
  xlab = "",
  ylab = "",
  main = "K-Means Clusters (k = 8)",
  asp = 1
)
plot(
  exdata3,
  pch = 16,
  col = pald_colors[exdata_hclust],
  xlab = "",
  ylab = "",
  main = "Hiearchical Clusters (k = 8)",
  asp = 1
)
```

@@

@@

This paper introduces the \CRANpkg{pald} package, demonstrating its utility for providing parameter-free clustering which can easily be implemented for a variety of data sets.
